---
title: "Analyzing Loan Data from Prosper"
author: "Yi Xu"
output:
  html_document:
    self_contained: no
---

# Overview of the data

Firstly, load in the necessary packages.
Then, read data into RStudio. We can see that the file has 113937 observations and 81 variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(dplyr)
library(ggplot2)
library(rMaps)
library(lubridate)
library(RColorBrewer)
library(GGally)
library(reshape2)
library(memisc)
library(knitr)
```

The names of the variables are also listed here.

```{r echo=FALSE, Load_the_Data}
prosper = tbl_df(read.csv("prosperLoanData.csv", stringsAsFactors = FALSE))
dim(prosper)
names(prosper)
```

Now, let's look at the structure of the dataset.

```{r data_structure, echo=FALSE}
str(prosper)
```

# Univariate Analysis

The first thinkg I would like to explore is the income range of the borrowers. We could see that most of the borrowers have income ranging from $25,000 to $75,000.

```{r IncomeRange, cache=TRUE, echo=FALSE}
prosper$IncomeRange = factor(prosper$IncomeRange, 
                             levels=c("Not employed", "$0", "$1-24,999", 
                                    "$25,000-49,999", "$50,000-74,999", 
                                    "$75,000-99,999", "$100,000+", 
                                    "Not displayed"))
ggplot(data = prosper, aes(x = IncomeRange, fill = IncomeRange)) + 
        geom_bar() + theme(axis.text.x  = element_text(angle=60, vjust=0.6)) +
        ggtitle("Income range of the borrowers")
```

Next I check how many of the incomes are verifiable.

```{r IncomeVerifiable, cache=TRUE, echo=FALSE}
prosper$IncomeVerifiable = as.factor(prosper$IncomeVerifiable)
ggplot(data = prosper, aes(x = IncomeVerifiable)) + 
        geom_bar(color = "black", fill = "cyan") +
        ggtitle("Income Verifiable?")
```

We can see that about 92.4% of incomes are verifiable. So we could trust the income range.

Another interesting variable is DebtToIncomeRatio, which shows the debt to income ratio when the borrower's profile was pulled.

```{r plot1, cache=TRUE,echo=FALSE}
ggplot(data = prosper, aes(x = DebtToIncomeRatio)) +                
        geom_histogram(color = "black", fill = "cyan", binwidth = 0.02) +
        xlab("Ratio") +
        xlim(0, quantile(prosper$DebtToIncomeRatio, prob = 0.99, na.rm=TRUE)) +
        ggtitle("Debt To Income Ratio")
```

We could see from the above histogram that most of the borrowers keep the ratio well below 50%.

Now I want to see the rating given by Prosper towards the buyers. There are two kinds of rating provided by Prosper: one is in the numeric form, the other is in the letter form. We could see from the following comparison that either form has 29084 missing values. So I just pick the letter form randomly. Note that here, rating "AA" is the best rating, while "HR" is the worst rating.

```{r ratings, cache=TRUE, echo=FALSE}
prosper$ProsperRating_n = prosper$ProsperRating..numeric.
prosper$ProsperRating_a = factor(prosper$ProsperRating..Alpha.,
                                 levels = c("AA","A","B","C","D","E","HR",""))
summary(prosper$ProsperRating_n)
table(prosper$ProsperRating_a)
```

We could see that most of the borrowers don't have ratings. This is because borrower rating was introduced in July 2009. For the borrowers who have ratings, most of them are rated as "A", "B", "C" or "D".

```{r plot2, cache=TRUE, echo=FALSE}
ggplot(data = prosper, aes(x = ProsperRating_a, fill = ProsperRating_a)) + 
        geom_bar() + theme(axis.text.x = element_text(angle = 60, vjust = 0.6)) +
        xlab("Rating") + ggtitle("Borrower Rating from Prosper")
```

Another variable similar to Prosper's borrower rating is Prosper's score towards each borrower, which is the risk built using historical Prosper data. The documentation says the score ranges from 1 to 10, with 10 being the best, or lowest risk score. However, there are the score ranges from 1 to 11 in the dataset.

```{r plot3, cache=TRUE,echo=FALSE}
prosper$ProsperScore = factor(prosper$ProsperScore)
ggplot(data = prosper, aes(x = ProsperScore, fill = ProsperScore)) + 
        geom_bar() + theme(axis.text.x = element_text(angle = 60, vjust = 0.6)) +
        xlab("Score") + ggtitle("Borrower Score from Prosper")
```

We can see that except for the missing values, borrowers' score are approximately normally distributed. Most of the borrowers have score ranging from 4 to 8. 

Since we also have borrower's credit score from credit rating angecies, let's investigate it.

```{r plot4, cache=TRUE, echo=FALSE}
ggplot(data = prosper, aes(x = CreditScoreRangeLower)) +
        geom_histogram(color = "black", fill = "yellow", binwidth = 20) + 
        xlab("Credit Score") +
        ggtitle("Borrower's Credit Score Lower Range")
ggplot(data = prosper, aes(x = CreditScoreRangeUpper)) +
        geom_histogram(color = "black", fill = "yellow", binwidth = 20) +
        xlab("Credit Score") +
        ggtitle("Borrower's Credit Score Upper Range")
```

Surprisingly, the lower and upper range of the borrower's credit score resemble each other. It appears to me that in order to get a Prosper loan, you'd better have a credit score over 620.

Next I would like to know the purpose of the loans. Therefore I investigate the listing category variable.

```{r Category, cache=TRUE,echo=FALSE, warning=FALSE}
str(prosper$ListingCategory..numeric.)
prosper$ListingCategory = factor(prosper$ListingCategory..numeric.,
                                 levels = c(1:6,8:20,7,0),
                                 labels = c('Debt Consolidation',
                                 'Home Improvement', 
                                 'Business', 
                                 'Personal Loan', 
                                 'Student Use', 
                                 'Auto', 
                                 'Baby & Adoption',
                                 'Boat', 
                                 'Cosmetic Procedure', 
                                 'Engagement Ring', 
                                 'Green Loans', 
                                 'Household Expenses', 
                                 'Large Purchases', 
                                 'Medical/Dental', 
                                 'Motorcycle', 'RV',
                                 'Taxes', 'Vacation',
                                 'Wedding Loans', 
                                 'Other', 
                                 'Not Available'))

ggplot(data = prosper, aes(x = ListingCategory, ymax = max(..count..)*1.05)) +
        geom_bar(fill="cyan") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.6)) + 
        stat_bin(geom="text", binwidth = 500, size = 3,
                 aes(label=..count.., vjust=-0.9, hjust=0.5)) +
        ggtitle("The Purpose of the Loan Listing")
```

From the above figure, we could tell that most loans on Prosper are for Debt Consolidation. Excluding the Not Available and Other purposes, most of the loans go into Home Improvement, Business, Auto, and Personal Loan. Categories like RV (recreatonal vehicle), Boat,  Vacation only account a small portion of the loan. Therefore, I conclude that most the people borrow money for necessary needs while not for entainment.

Since borrowers' interest rate is an important factor, let's take a look at this variable.

```{r plot5, cache=TRUE,echo=FALSE}
ggplot(data = prosper, aes(x = BorrowerRate)) + 
        geom_histogram(color = "black", fill = "lightblue", binwidth = 0.01) +
        xlab("Rate") +
        ggtitle("Borrowers' Rate Distribution")
```
We could see that the rate approximately follows a normal distribution with mean around 0.18.

The length of a loan is also an important factor for both the borrower and investor. So let's take a look at this variable.

```{r Term, cache=TRUE,echo=FALSE}
prosper$Term = as.factor(prosper$Term)
ggplot(data = prosper, aes(x = Term, ymax = max(..count..)*1.05)) + 
        geom_bar(fill = "cyan") +
        stat_bin(geom="text", size = 4,
                 aes(label=..count.., vjust=-0.9, hjust=0.5))+
        ggtitle("Length of the Loan")
```

We could see from this barplot that the majority of the loans has the length of 36 months. About 20% of loans have length of 60 months. Only 1.4% of the loans have length of 12 months.

Next, I want to investigate the delinquent amount of the borrower.

```{r plot6,cache=TRUE,echo=FALSE, warning=FALSE}
ggplot(data = prosper, aes(x = AmountDelinquent)) +
        geom_histogram(color = "black", fill = "cyan", binwidth = 100) +
        xlim(0, quantile(prosper$AmountDelinquent, 0.95, na.rm=TRUE)) +
        xlab("Amount in Dollars") +
        ggtitle("Delinquent Amount")
```

We can see that the delinquent amount for almost all the borrowers is 0. 

Now, let's look at the borrower's geographical informaiton. I first count the number of borrowers in each state, then color the map according to this number.

```{r State,cache=TRUE,echo=FALSE}
prosper$BorrowerState = as.factor(prosper$BorrowerState)
```


```{r StateMap, cache=TRUE,echo=FALSE}
borrower_state_info = prosper %>%
        group_by(BorrowerState) %>%
        summarise(n = n()) %>%
        arrange(desc(n))
```

```{r statemap1, results='asis', cache=TRUE,echo=FALSE}
borrower_state_map <- ichoropleth(n ~ BorrowerState, data = borrower_state_info,
                                  ncuts = 5, pal = 'PuRd', 
                                  geographyConfig = list(
                                          popupTemplate = "#!function(geo, data) {
                                          return '<div class=\"hoverinfo\"><strong>' +
                                          data.BorrowerState +
                                          '<br>' + data.n +
                                          '</strong></div>';}!#"))
print(borrower_state_map)
```

As we can see from this map, California, Texas, Georgia, Florida, and New York have the most borrowers. These are the most populous and economically prosperous areas as well.

Recommendation is also an important factor for loans. 

```{r plot7, cache=TRUE, echo=FALSE}
ggplot(data = prosper, aes(x = Recommendations)) + 
        geom_histogram(binwidth = 1, color = "black", fill = "cyan") +        
        xlab("Recommendation Count") +
        ggtitle("Number of Recommendations of Loan")
```
We can see that most loans do not have any recommendations.

We may also take a look at the number of investors of each loan.

```{r Investors, cache=TRUE, echo=FALSE}
ggplot(data = prosper, aes(x = Investors)) + 
        geom_histogram(binwidth = 50, color = "black", fill = "cyan") +        
        ggtitle("Number of Investors of Loan")
```

We can see that this data right skewed. Most of the loans have less than 50 investors.



Finally, let's look at the lenders' profit, which is contained in the variable LenderYield.

```{r plot8, cache=TRUE, echo=FALSE}
ggplot(data = prosper, aes(x = LenderYield)) +
        geom_histogram(color = "black", fill = "cyan", binwidth = 0.01) + 
        xlab("Yield") +
        ggtitle("Lenders' Yield")
```

Comparing this histogram to the Borrowers' rate histogram, I find them to be nearly identical.


# Bivariate Analysis


I first look into the relationship between the length of a loan and the yield of the lender.

```{r plot9, echo=FALSE, cache=TRUE}
ggplot(data = prosper, aes(x = LenderYield)) +
        geom_histogram(color = "black", fill = "cyan",binwidth = 0.01) + 
        facet_wrap(~Term, ncol = 1) +
        xlab("Yield") +
        ggtitle("Lender's Yield of Different Terms")
```

We could see that most of the loads are for 36 months. And 36 months also have very high return, such as 0.3 compared to other two terms. Given that 36-month is much shorter than 60-month, and 36-month even has higher return rate, I would suggest the lenders putting their money into 36-month loans.

Next, I look at the borrower's rate with different income range.

```{r plot10, echo=FALSE, cache=TRUE}
ggplot(data = prosper, aes(x = BorrowerRate, fill = IncomeRange)) +
        geom_histogram(binwidth = 0.01) +         
        xlab("Rate") + 
        ggtitle("Borrower's Rate of Different Income Range")
```

We could see that within each income range, the rate seems to be normally distributed. About half of the borrowers in each income range have rate less than 0.2, and the other half greater than 0.2.

Now I check the relationship between the delinquent amount and income range.

```{r plot11, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = AmountDelinquent, fill = IncomeRange)) +
        geom_density() +         
        xlab("Amount in Dollars") + 
        facet_wrap(~IncomeRange, ncol = 2) +
        coord_cartesian(xlim = c(0, quantile(prosper$AmountDelinquent, 0.99, na.rm = TRUE))) +
        ggtitle("Delinquent Amount with Different Income Range")
```

We could see that among different income ranges, there are no big differences. All of income levels have small delinquent amount.

Let's see how the number of investors and the lender yield are correlated.

```{r plot12, echo=FALSE, cache=TRUE}
ggplot(data = prosper, aes(x = Investors, y = LenderYield, alpha = 0.01)) +
        geom_jitter() +
        ggtitle("Lender Yield Vs Number of Investors")
```

According to the above scatter plot, it seems that there is no strong correlation between these two variables. We could calculate their correlations as follows.

```{r correlation, echo=FALSE, cache=TRUE}
with(prosper, cor(Investors, LenderYield))
```

The result is -0.2741739, which indicates that the correlation is not very strong.

Does different states have different yields?

```{r statemap2, results='asis', cache=TRUE, echo=FALSE, fig.keep='all'}
lenderyield_state_info = prosper %>%
        group_by(BorrowerState) %>%
        summarise(mean_yield = mean(LenderYield)) 

lenderyield_state_map = 
    ichoropleth(mean_yield ~ BorrowerState, data = lenderyield_state_info,
                ncuts = 5, pal = 'Greens', 
                geographyConfig = list(
                        popupTemplate = "#!function(geo, data) {
                        return '<div class=\"hoverinfo\"><strong>' +
                        data.BorrowerState +
                        '<br>' + data.mean_yield +
                        '</strong></div>';}!#"))

print(lenderyield_state_map)
```

From the above plot, we can see that states in the south, e.g., Mississippi, Alabama, Arkansas, Tennessee have very high yields on average. States in the middle, for instance, Colorado, Oklahoma, Kansas, Nebraska have slightly lower yield.


Now let's look at the relationship between lender yield and the score from Prosper.
In common sense, investors want to lend their money to someone trustable. 

```{r plot13, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = ProsperScore, y = LenderYield)) +
        geom_boxplot(aes(fill = ProsperScore)) +
        xlab("Score") +
        ggtitle("Yield for Different Prosper Score")
```

From the boxplot above, we could see that, except for some outliers, lending money to the borrower of low Prosper score seems to yield higher return. To confirm this, let's look at the relationship between the Prosper rating and lender yield.

```{r plot14, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = ProsperRating_a, y = LenderYield)) +
        geom_boxplot(aes(fill = ProsperRating_a)) +
        xlab("Prosper Rating") +
        ggtitle("Yield for Different Prosper Rating")
```

Since AA is the highest rating and HR is lowest rating, the same conclusion holds. That is lending money to low rating borrowers will produce higher returns. This is very interesting and kind of counter-intuitive. However, after careful thought, I think it also makes sense. Since when Prosper lend money to the lowe rating borrowers, they will set higher APR to compensate for the high risk of not getting money back. The following plot shows the different APR for different rating borrowers.


```{r plot15, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = ProsperRating_a, y = BorrowerAPR)) +
        geom_boxplot(aes(fill = ProsperRating_a)) +
        xlab("Prosper Rating") +
        ggtitle("APR for Different Prosper Rating")
```

Therefore, we could conclude that lending money to low rating borrower could give us higher return.

How about different listing categories?

```{r plot16, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = ListingCategory, y = LenderYield)) +
        geom_boxplot(aes(fill = ListingCategory)) +
        xlab("Category") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.6)) +
        ggtitle("Lender Yield for Different Listing Category")
```

From this boxplot, it seems that there is not clear relationship between these two variables.
Can the delinquent amount serve as a good indicator for lender yield?

```{r plot17, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = AmountDelinquent, y = LenderYield, alpha = 0.01)) +
        geom_jitter() +
        xlab("Amount") +
        ggtitle("Lender Yield Vs Delinquent Amount")
```

Since there are not many borrowers with delinquent amount larger than 0, it seems that there is a weak relationship between these two variables.

How about the number of recommendations? Does more recommendations usually result in higher return?

```{r plot18, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = Recommendations, y = LenderYield, alpha = 0.01)) +
        geom_jitter() +
        xlab("Number of Recommendations") +
        ggtitle("Lender Yield Vs Number of Recommendations")
```

From this scatter plot, it appears to me that more recommendations do not always have high return.

Lastly, how about the borrower's debt to income ratio? Can we lend our money to someone with high debt to income ratio?

```{r plot19, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = DebtToIncomeRatio, y = LenderYield, alpha = 0.01)) +
        geom_jitter() +
        xlab("Ratio") +
        ggtitle("Lender Yield Vs Borrower's Debt to Income Ratio")
```

We could that the bulk of the borrowers has debt to income ratio less than 1. And it seems that higher ratio has higher return.


# Multivariate Analysis

Firstly, let's see if the number of investor for different income range borrowers have impacts on the returns?

```{r plot20, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = prosper, aes(x = Investors, y = LenderYield, color = IncomeRange)) +
        geom_jitter() +
        xlab("Number of Investors") +
        ggtitle("Lender Yield Vs Number of Investors")
```

From this plot, I didn't see anything interesting ongoing on.

```{r plot21, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = subset(prosper, ProsperRating_a != ""), 
       aes(x = AmountDelinquent, y = LenderYield, color = ProsperRating_a)) +
        geom_jitter() +
        xlab("Amount") +
        ggtitle("Lender Yield Vs Delinquent Amount")
```

From the above plot, we could see that lower rating borrower and borrower with high delinquent amount seem to have higher return. How about debt to income ratio?

```{r plot22, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = subset(prosper, ProsperRating_a != ""), 
       aes(x = DebtToIncomeRatio, y = LenderYield, color = ProsperRating_a)) +
        geom_jitter() +
        xlab("Ratio") +
        ggtitle("Lender Yield Vs Debt to Income Ratio")
```

The above plot shows that there is a strong relationship between lender yield and debt to income ratio. If I take log of the lender yield, the relationship seems to be more clear.

```{r plot23, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = subset(prosper, ProsperRating_a != ""), 
       aes(x = DebtToIncomeRatio, y = LenderYield, color = ProsperRating_a)) +
        geom_jitter() +
        scale_y_log10() +
        xlab("Ratio") +
        ylab("Log Lender Yield") +
        ggtitle("Lender Yield Vs Debt to Income Ratio")
```


Are there some interactions between the length of loans and the Prosper rating for yield?

```{r plot24, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = subset(prosper, ProsperRating_a != ""), 
       aes(x = ProsperRating_a, y = LenderYield)) +
        geom_boxplot(aes(fill = ProsperRating_a)) +
        facet_wrap(~Term) +
        xlab("Prosper Rating") +
        ggtitle("Yield for Different Prosper Rating")
```

From this plot, we could tell that for each rating, 60-month loan tends to have the highest average return, while 12-month load tends to have lowest average return.


```{r model, cache=TRUE, echo=TRUE}
m1 = lm(LenderYield ~ DebtToIncomeRatio, data = prosper)
m2 = update(m1, ~. + AmountDelinquent)
m3 = update(m2, ~. + ProsperRating_a)
m4 = update(m3, ~. + Term)
m5 = update(m4, ~. + BorrowerState)
mtable(m1, m2, m3, m4, m5)
```


# Final Plots and Summary

## Plot One

```{r plot25, cache=TRUE, echo=FALSE}
ggplot(data = prosper, aes(x = LenderYield)) +
        geom_histogram(color = "black", fill = "cyan", binwidth = 0.01) + 
        xlab("Yield") +
        ggtitle("Lenders' Yield")
```

## Description One

The lender's yield approximated follows a normal distribution, with center around 0.15. There is a spike around 0.3.

## Plot Two

```{r plot26, results = 'asis', echo=FALSE, cache=TRUE}
print(lenderyield_state_map)
```

````{r plot262, echo=FALSE, cache=TRUE}
ggplot(data = prosper, aes(x = ProsperRating_a, y = LenderYield)) +
        geom_boxplot(aes(fill = ProsperRating_a)) +
        xlab("Prosper Rating") +
        ggtitle("Yield for Different Prosper Rating")
```

## Description Two

From the above two plots, I made two observations. The first is that different states have different level of returns. States like Mississippi, Alabama, Arkansas, Tennessee tend to have high yields on average. Secondly, borrower rating from Prosper can be used to predict the result of the loan.


## Plot Three

```{r plot27, echo=FALSE, cache=TRUE, warning=FALSE}
ggplot(data = subset(prosper, ProsperRating_a != ""), 
       aes(x = DebtToIncomeRatio, y = LenderYield, color = ProsperRating_a)) +
        geom_jitter() +
        scale_y_log10() +
        xlab("Ratio") +
        ylab("Log Lender Yield") +
        ggtitle("Lender Yield Vs Debt to Income Ratio")

ggplot(data = subset(prosper, ProsperRating_a != ""), 
       aes(x = ProsperRating_a, y = LenderYield)) +
        geom_boxplot(aes(fill = ProsperRating_a)) +
        facet_wrap(~Term) +
        xlab("Prosper Rating") +
        ggtitle("Yield for Different Prosper Rating")
```

## Description Three

The debt to income ratio can be served as an indicator for the yield of the loan. Although for low debt to income ratio borrowers it is hard to predict the yield, for high debt to income ratio borrowers, lending money to money usually have high return. Also, different lengths of loan tend to different yields. From the plot, we can see that on average, lending money to borrower with Prosper rating HR for 36 month will have highest return.






# Reflections

The Prosper loan data set has 113,937 transaction record with 81 variables. I explored 15 out of these 81 variables. I started by looking at the documentation and tried to find interesting variables. Then I used various plots to check how these variable distributed. 
I struggled with understanding this dataset and asking interesting questions out of this dataset. To further understand this dataset, I tried to visualize the interaction between two variables. For instance, I investigated how BorrowerRate is correlated with LenderYield, etc. Then one question came into my mind: if I were a lender, what kind of borrower should I lend my money to? 
From the lender's perspective, he doesn't care about the borrower's income range, the borrower's credit score, the borrower's purpose of borrowing money or the borrower's delinquent amount.
The only thing he cares about is to make more money. With this on mind, I made more bivariate plots and tried to find which variables are correlated with the LenderYield variable. 
I found that the rating from Prosper was an good indicator in a seemingly strange way. That is the lower the rating is, the higher the expected yield is. This is surprising at first. Then I found that this was true, since we need higher return to compensate high risk. I also found the BorrowerState, Term, AmountDelinquent, DebtToIncomeRatio helpful to predicting the loan result through multivariate analysis. To summarize my finding in one single sentence, you should lend your money to some borrowers from Alabama, with Propser rating HR, high delinquent amount, high debt to income ratio for 36 or 60 months.


#Reference
1. Prosper official site: https://www.prosper.com/
2. rMaps: http://www.r-bloggers.com/rmaps-mexico-map/
3. R Colors: http://vis.supstat.com/2013/04/plotting-symbols-and-color-palettes/
4. Tips for rMaps and rCharts: http://daisukeichikawa.blogspot.com/2014/03/tips-for-rmaps-and-rcharts.html